An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-05-30 demo.tex
pre methods update

@@

\title[Short Title]{An interdisciplinary examination of stress and injury occurrence in
athletes}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

@@

A challenge when adopting a complex systems approach is using an
appropriate analysis technique that is able to capture the uncertainty
and complexity of the relationships between different factors. One
technique that provides a solution to this challenge is Bayesian network
(BN) modelling. BN's allow the construction of graphical probabilistic
models using the underlying structure that connects different variables
(nodes in the network) of interest (Scutari and Denis, 2014). Once
constructed, the learned structure can be used for inference by
obtaining the posterior probabilities of a particular node for a given
query (e.g., if the value of Node A is x and the value of Node B is y,
what is the probability Node C being value z?). Furthermore, BN's do not
distinguish between dependent and independent variables as they are a
form of unsupervised learning, which is a strength over regression or
structure equation models when the underlying relationship in the
network may not be known (Olmedilla et al., 2018).

To summarise, despite offering a possible framework to build on the
research stemming from Williams and Andersen's (Williams and Andersen,
1998) model, there remains and opportunity to explore other
physiological stress-related markers proposed by the BMSAIH, in addition
to the already well-established psychological characteristics known to
be related to injury (Appaneal and Perna, 2014). Furthermore, research
has typically not captured changes in both psychosocial factors and
stress-related physiological markers that may occur between initial
measurement and injury occurrence. Given the exploratory the current
study had the following objectives:

\begin{itemize}
\tightlist
\item
  Identify suitable markers of stress that can be easily captured in a
  large cohort of athletes in a field based setting.
\item
  Examine the relationships between the markers of stress and injury
  using a prospective, repeated measures design.
\item
  Evaluate the relationships between the markers of stress and injury
  using Bayesian network modelling that captures the complex nature of
  injury occurrence.
\end{itemize}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Andersen1988}{}%
Andersen, M. B., and Williams, J. M. (1988). A model of stress and
athletic injury: Prediction and prevention. \emph{Journal of Sport and
Exercise Psychology} 10, 294--306.
doi:\href{https://doi.org/10.1123/jsep.10.3.294}{10.1123/jsep.10.3.294}.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\title[Stress and sports injury]{An interdisciplinary examination of stress and injury occurrence in
athletes}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

@@

A challenge when adopting a complex systems approach is using an
appropriate analysis technique that is able to capture the uncertainty
and complexity of the relationships between different variables. One
technique that provides a solution to this challenge is Bayesian network
(BN) modelling. BN's allow the construction of graphical probabilistic
models using the underlying structure that connects different variables
(nodes in the network) (Scutari and Denis, 2014). The learned structure
can be used for inference by obtaining the posterior probabilities of a
particular node for a given query (e.g., if the value of Node A is x and
the value of Node B is y, what is the probability Node C being value
z?). Furthermore, BN's do not distinguish between dependent and
independent variables as they are a form of unsupervised learning, which
is a strength over regression or structure equation models when the
underlying relationship in the network may not be known (Olmedilla et
al., 2018).

To summarise, despite offering a possible framework to build on the
research stemming from Williams and Andersen's (Williams and Andersen,
1998) model, there remains and opportunity to explore other
physiological stress-related markers proposed by the BMSAIH, in addition
to the already well-established psychological characteristics known to
be related to injury (Appaneal and Perna, 2014). Furthermore, research
has typically not captured changes in both psychosocial factors and
stress-related physiological markers that may occur between initial
measurement and injury occurrence. Given the exploratory nature of the
current study the following objectives were defined:

\begin{itemize}
\tightlist
\item
  Identify suitable markers of stress that can be easily captured in a
  large cohort of athletes in a field based setting.
\item
  Capture the markers of stress and injury occurrence using a
  prospective, repeated measures design.
\item
  Explore and evaluate the relationships between the markers of stress
  and injury using Bayesian network modelling.
\end{itemize}

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{participants}{%
\subsection*{Participants}\label{participants}}
\addcontentsline{toc}{subsection}{Participants}

A total of 351 athletes (male: \emph{n} = 231, female: \emph{n} = 120)
were initially recruited for the study. Participants had an average age
of 22.0 \(\pm\) 7.0 years and represented a range of team (football,
rugby, netball, cricket, lacrosse, basketball and field hockey) and
individual sports (athletics, tennis, weightlifting, gymnastics, judo,
swimming and golf from a British University and local sports clubs
(Table \ref{tab:table1}). Participants self-rated competitive level
ranged from recreational to international standard. A total of 126
(49.03\%) participants had sustained an injury in the 12 months prior to
the start of the study (male: \emph{n} = 114 {[}49\%{]}, female;
\emph{n} = 48 {[}40\%{]}). At the start of the study, all participants
were injury free (no modifications to their usual training routine due
to a sport related medical problem for a minimum of four weeks).
Participants were engaged in training for their respective sports for at
least five hours per week. Ethical approval was obtained from the
University ethics committee prior to the start of the study and all
participants provided informed consent.

\begin{table}[H]

\caption{\label{tab:table1}Participant characteristics.}
\centering
\begin{tabular}[t]{c|c|c}
\hline
\textbf{ } & \textbf{Female (n = 120)} & \textbf{Male (n = 231)}\\
\hline
\multicolumn{3}{l}{\textbf{Demographics M (SD)}}\\
\hline
\hspace{1em}Age (yrs) & 26.0 (11.3) & 20.2 (1.8)\\
\hline
\hspace{1em}Height (cm) & 167.4 (7.6) & 177.8 (7.8)\\
\hline
\hspace{1em}Body mass (kg) & 67.0 (9.5) & 82.0 (14.6)\\
\hline
\hspace{1em}Hours per week training & 8.5 (4.5) & 11.2 (8.8)\\
\hline
\multicolumn{3}{l}{\textbf{Current competitive level n (\%)}}\\
\hline
\hspace{1em}Recreational & 3 (4) & 7 (4)\\
\hline
\hspace{1em}University & 45 (56) & 141 (80)\\
\hline
\hspace{1em}National/International & 33 (41) & 28 (16)\\
\hline
\end{tabular}
\end{table}

\hypertarget{measures}{%
\subsection*{Measures}\label{measures}}
\addcontentsline{toc}{subsection}{Measures}

\hypertarget{major-life-events}{%
\subsubsection*{Major life events}\label{major-life-events}}
\addcontentsline{toc}{subsubsection}{Major life events}

A modified version of the Life Events Survey for Collegiate Athletes
(LESCA) was used to measure participants' history of life event stress
(Petrie, 1992). Modifications were made to the LESCA to ensure the
suitability of the items for the study population (S1 Table). The LESCA
comprises 69 items that reflect possible life events that participants
may have experienced. Example items include, ``Major change in the
frequency (increased or decreased) of social activities due to
participation in sport'', ``Major change in the amount (more or less) of
academic activity (homework, class time, etc)'' and ``Major change in
level of athletic performance in actual competition (better or worse)''.
Participants were asked to rate the perceived impact of each life event
they had experienced within the last 12 months on an 8-point Likert
scale anchored at -4 (\(extremely\;negative\)) and +4
(\(extremely\;positive\)). Negative and positive life event scores were
calculated by summing the negative and positive scores respectively. A
score for total life events was calculated by summing the absolute
values for both negative and positive events. Petrie (1992) (Petrie,
1992) reported test-retest reliabilities at 1-week and 8-weeks with
values ranging from .76 to .84 (p \textless{} .001) and .48 to .72 (p
\textless{} .001) respectively. Petrie also provided evidence of
discriminant, convergent and predictive validity. The LESCA is the most
widely used measure of major life events for athletes in the sports
injury literature. For this study, Composite Reliability (Fornell and
Larcker, 1981) was preferred to Cronbach's alpha as it does not assume
parallelity (i.e., all factor loadings are constrained to be equal, and
all error variances are constrained to be equal) and instead takes into
consideration the varying factor loadings of the items in the
questionnaire. The composite reliability for the LESCA in this study was
0.84.

\hypertarget{reinforcement-sensitivity-theory-personality-questionnaire}{%
\subsubsection*{Reinforcement Sensitivity Theory Personality
Questionnaire}\label{reinforcement-sensitivity-theory-personality-questionnaire}}
\addcontentsline{toc}{subsubsection}{Reinforcement Sensitivity Theory
Personality Questionnaire}

A revised version of the Reinforcement Sensitivity Theory Personality
Questionnaire (RST-PQ) was used to measure motivation, emotion,
personality and their relevance to psychopathology (Corr and Cooper,
2016). The revised version of the RST-PQ comprises 51 statements that
measure three major systems: Fight-Flight-Freeze System (FFFS; e.g., ``I
am the sort of person who easily freezes-up when scared''), Behavioural
Inhibition System (BIS; e.g., ``When trying to make a decision, I find
myself constantly chewing it over'') and four Behavioural Approach
System (BAS) factors; Reward Interest (e.g., ``I regularly try new
activities just to see if I enjoy them''), Goal Drive Persistence (e.g.,
``I am very persistent in achieving my goals''), Reward Reactivity
(e.g., ``I get a special thrill when I am praised for something I've
done well'') and Impulsivity (e.g., ``I find myself doing things on the
spur of the moment''). Participants rated each item on a scale from 1
(\(not\; at\; all\)) to 4 (\(highly\)) to reflect how well each
statement described their personality in general. The responses to items
associated with each subscale (FFFS, BIS, RI, GDP, RR and I) were summed
to give a total score that was subsequently used for further analysis.
The composite reliabilities for each subscale were; BIS = 0.92, FFFS =
0.77, GDP = 0.87, I = 0.71, RI = 0.77, RR = 0.81. Further details
regarding the revised RST are in S1 Appendix.

\hypertarget{heart-rate-variability}{%
\subsubsection*{Heart rate variability}\label{heart-rate-variability}}
\addcontentsline{toc}{subsubsection}{Heart rate variability}

A Polar V800 heart rate monitor (HRM) and Polar H7 Bluetooth chest strap
(Polar OY, Finland) was used to collect inter-beat interval (IBI) data.
IBI recordings using the Polar V800 are highly comparable (ICC = 1.00)
with ECG recordings (Giles et al., 2016), which are considered the gold
standard for assessing HRV. In addition, HRV indices calculated from IBI
and ECG data have shown a strong correlation (r = .99) in athletes
(Caminal et al., 2018) and under spontaneous breathing conditions (Plews
et al., 2017).

\hypertarget{musculoskeletal-properties}{%
\subsubsection*{Musculoskeletal
properties}\label{musculoskeletal-properties}}
\addcontentsline{toc}{subsubsection}{Musculoskeletal properties}

A handheld myometer (MyotonPRO, Myoton AS, Tallinn, Estonia) was used to
measure muscle stiffness. The MyotonPRO is a non-invasive, handheld
device that applies a mechanical impulse of 0.40 N for 0.15 ms
perpendicular to the surface of the skin. The impulse causes natural
damped oscillations in the tissue, which are recorded by a three-axis
digital accelerometer sensor in the device. The raw oscillation signal
is then processed, and the stiffness parameter is calculated
(Agyapong-Badu et al., 2016). The MyotonPRO has previously been reported
to be a reliable and valid tool for the measurement of in-vivo tissue
stiffness properties (Chuang et al., 2013; Pruyn et al., 2016; Nair et
al., 2014), and has demonstrated good internal consistency (coefficient
of variation \textless{} 1.4\%) over sets of 10 repetitions (Aird et
al., 2012).

\hypertarget{postural-stability}{%
\subsubsection*{Postural stability}\label{postural-stability}}
\addcontentsline{toc}{subsubsection}{Postural stability}

Postural stability was assessed with a modified version of the balance
error scoring system (mBESS) based on the protocol recommended by Hunt
et al. (2009). In total, each trial of the mBESS was performed without
shoes (McCrory et al., 2013) and included six stances in the following
order; dominant leg (DL; standing on the dominant foot with the
non-dominant foot at approximately 30-degrees of hip flexion and
45-degrees of knee flexion), non-dominant leg (NDL; standing on the
non-dominant foot with the dominant foot at approximately 30-degrees of
hip flexion and 45-degrees of knee flexion) and tandem leg stance (TS;
standing heel-to-toe with the non-dominant foot behind the dominant) on
firm and foam (Alcan airex AG, Sins, Switzerland) surfaces respectively
(Fig \ref{fig:fig2}). To determine leg dominance, participants were
asked their preferred leg to kick a ball to a target, and the chosen
limb was labelled as dominant (cf.~Cingel et al., 2017). Participants
were asked to maintain each stance for a total of 20-seconds.
Participants hands were placed on hips at the level of the iliac crests.
A Sony DSC-RX10 video camera (Sony Europe Limited, Surrey, United
Kingdom) was used to record each participants performance during the
mBESS.

\begin{figure}[!h]
\caption{{\bf mBESS positions (A-F).}
Top row, firm surface. Bottom row, foam surface. Left column, dominant leg stance. Middle column, non-dominant leg stance. Right column, Tandem leg stance.}
\label{fig:fig2}
\end{figure}

The error identification criteria from the original BESS protocol was
used by the lead researcher who scored all the BESS trials. One error
was recorded if any of the following movements were observed during each
trial: a) lifting hands off iliac crests; b) opening eyes; c) stepping,
stumbling, or falling; d) moving the thigh into more than 30 degrees of
flexion or abduction; e) lifting the forefoot or heel; and f) remaining
out of the testing position for more than 5-seconds (Riemann et al.,
1999). A maximum score of 10 errors was possible for each stance.
Multiple errors occurring simultaneously were recorded as one error. A
participant was given the maximum score of 10 if they remained out of
the stance position for more than 5-seconds. To calculate limb
asymmetry, the DL and NDL leg score was calculated by summing the DL and
NDL errors respectively. A total score was calculated by summing the
total number of errors recorded on all stances (DL, NLD and TS, on foam
and firm surfaces). To assess the intra-rater reliability, a single
measurement, absolute agreement, two-way mixed effects model for the
intraclass correlation (ICC; Koo and Li, 2016) was used on a sample of
40 participants from the first time point. The test-retest scoring of
BESS resulted in a ``good'' to ``excellent'' ICC score (ICC = 0.93, 95\%
confidence interval = 0.88 - 0.96), indicating the scoring was reliable.

\hypertarget{injury}{%
\subsubsection*{Injury}\label{injury}}
\addcontentsline{toc}{subsubsection}{Injury}

Participants self-reported any injuries they sustained at each data
collection during the study period. An injury was defined as any sports
related medical problem causing the athlete to miss or modify their
usual training routine for at least 24 hours (Fuller et al., 2006, 2007;
Timpka et al., 2014). Minor scrapes and bruises that may require certain
modifications (e.g., strapping or taping) but did not limit continued
participation were not considered injuries (cf.~Appaneal et al., 2009).
Injury status (did / did not sustain an injury) served as the main
outcome measure.

\hypertarget{procedure}{%
\subsection*{Procedure}\label{procedure}}
\addcontentsline{toc}{subsection}{Procedure}

At the start of the UK academic year (September 2016 and 2017), coaches
of sports teams at a British University and local sports clubs were
contacted and informed about the study. With the coaches' permission,
the lead researcher attended training sessions to inform the athletes
about the overall purpose of the study and the requirements of
participation. To be eligible for the study athletes had to be injury
free (no modifications to their usual training routine due to a sport
related medical problem for a minimum of four weeks) and training a
minimum of five hours per week. Athletes who met the criteria and
volunteered to take part in the study were invited to attend scheduled
testing sessions. A repeated measures prospective cohort design was used
to assess athletes' major life events, stress-related physiological
markers and injury status over two consecutive twelve-month periods
between September 2016 and September 2018. Each participant was asked to
attend a total of 3 data collections over a twelve-month period, with
each data collection separated by a four-month interval (Fig
\ref{fig:fig3}). Participants provided informed consent before data
collection commenced.

\begin{figure}[!h]
\caption{{\bf Study design.}
For each time point (T), each box contains the number of participants who completed the data collection (N), the measures used for data collection and the approximate date of the data collection.}
\label{fig:fig3}
\end{figure}

For the first three data collections (T1, T2 and T3), participants
followed the same protocol in a specific order (Fig \ref{fig:fig4}). To
ensure all measures could be collected within a viable time-frame,
participants were separated into two groups. The first group completed
all computer-based measures followed by all physical measurements,
whereas the second group completed all physical measurements followed by
computer-based measures. Participants were randomly assigned to one of
the two groups and remained in those groups across all time points.

\begin{figure}[!h]
\caption{{\bf Session protocol.}
Outline of the protocol for each data collection.}
\label{fig:fig4}
\end{figure}

\hypertarget{questionnaires}{%
\subsubsection*{Questionnaires}\label{questionnaires}}
\addcontentsline{toc}{subsubsection}{Questionnaires}

The questionnaires, which included demographic information, the LESCA,
RST-PQ (T1, T2, T3) and injury status (T2, T3, T4) were completed
on-line (SurveyMonkey Inc., USA, www.surveymonkey.com). The instructions
for the LESCA were modified at T2 and T3 so that participants reported
major life events that had occurred since the previous testing session.
For injury reporting, participants were asked to record any injuries
that they had sustained since the last data collection. The data were
downloaded from surverymonkey.com and imported into R (R Core Team,
2019) for analysis purposes.

\hypertarget{hrv}{%
\subsubsection*{HRV}\label{hrv}}
\addcontentsline{toc}{subsubsection}{HRV}

To minimise potential distractions, participants were directed to a
designated quiet area in the laboratory where IBI data were recorded.
Participants were instructed to turn off their mobile devices to avoid
any interference with the Bluetooth sensor. Each chest strap was
dampened with water and adjusted so it fitted tightly but comfortably,
as outlined by Polar's guidelines. Participants were seated and asked to
remain as still as possible for the duration of the recording. No
attempt was made to control participants respiratory frequency or tidal
volume (Denver et al., 2007). Inter-beat interval (IBI) data was
collected for 10-minutes at a sampling frequency of 1000 Hz.

Raw, unfiltered IBI recordings were exported from the Polar Flow web
service as a space delimited .txt file and imported into R (R Core Team,
2019) where the \emph{RHRV} package ({\textbf{???}}) was used to
calculate HRV indices. Raw IBI data was filtered using an adaptive
threshold filter, and the first 3-minutes and last 2-minutes of each
recording were discarded, leaving a 5-minutes window that was used to
calculate the root mean square of successive differences (RMSSD) in RR
intervals following the recommendations for short term IBI recordings
(Laborde et al., 2017; Malik et al., 1996). RMSSD was calculated as:

\begin{equation} 
  \overline{RR} = \frac{1}{N} \sum_{i=1}^{n} RR_i
\end{equation}

Where N is the length of the time series, and \(RR_i\) the RR interval
between beats \(i\) and \(i-1\), where each beat position corresponds to
the beat detection instant.

\hypertarget{muscle-stiffness}{%
\subsubsection*{Muscle stiffness}\label{muscle-stiffness}}
\addcontentsline{toc}{subsubsection}{Muscle stiffness}

To assess muscle stiffness, participants lay horizontally on a massage
bed and four testing sites were identified on each lower limb. The
muscle belly of the rectus femoris (RF), biceps femoris (BF), medial
gastrocnemius (MG) and lateral gastrocnemius (LG) sites were identified
using a visual-palpatory technique to determine the exact location of
each site (Chuang et al., 2012). The visual-palpatory technique required
the participant to contract the target muscle to aid the lead researcher
to visually identify the muscle. The participant was then asked to relax
the muscle and the muscle was palpated to locate the muscle belly. A
skin safe pen (Viscot all skin marker pen, Viscot Medical LLC, NJ) was
used to mark the testing site in the centre of the muscle belly.

After the eight testing sites had been identified, the testing end of
the MyotonPRO (diameter = 3 mm) was positioned perpendicular to the skin
on the testing site. A constant pre-load of 0.18 N was applied for
initial compression of subcutaneous tissues. The device was programmed
to deliver five consecutive impulses, separated by a one second interval
(Morgan et al., 2018). For each impulse, the device computed stiffness
values, with the median of the five values being saved by the device for
further analysis. In accordance with Myoton.com, a set of five
measurements with a coefficient of variation (CV) of less than 3\% was
accepted. Sets of measurements above 3 \% were measured again to ensure
the reliability of the data. The CV was calculated in real time by the
device after each set of measurements. Measurements saved on the device
were uploaded to a computer using MyotonPRO software and imported in R
(R Core Team, 2019) for further analysis. For each participant, the sum
of all eight testing sites was calculated to provide a total lower
extremity stiffness score and was used for further analysis.

\hypertarget{postural-stability-1}{%
\subsubsection*{Postural stability}\label{postural-stability-1}}
\addcontentsline{toc}{subsubsection}{Postural stability}

Instructions for the mBESS were read to each participant and a
demonstration of the positions was provided by the research assistant.
For each position, participants were instructed to close their eyes,
rest their hands on their iliac crests and remain as still as possible
for 20-seconds. Participants were instructed to return to the testing
position as quickly as possible if they lost their balance. The video
recording was started prior to the first stance position and stopped
after all stances had been completed. Each completed mBESS protocol took
approximately 4 minutes. Only one trial was performed to avoid
familiarisation effects across the repeated measurement (cf.~Valovich et
al., 2003). The video recordings for each participant were imported from
the recording equipment (Sony DSC-RX10) and the lead researcher scored
each trial using the error identification criteria.

\hypertarget{data-analysis}{%
\subsection*{Data Analysis}\label{data-analysis}}
\addcontentsline{toc}{subsection}{Data Analysis}

Two Bayesian Networks (BN) were used to explore the relationships
between the psychological measures, physiological markers of stress and
sports injury. A BN is a graphical representation of a joint probability
distribution among a set of random variables, and provides a statistical
model describing the dependencies and conditional independences from
empirical data in a visually appealing way (Scutari and Denis, 2014). A
BN consists of arcs and nodes that together are formally known as a
directed acyclic graph (DAG), where a node is termed a parent of a child
if there is an arc directed from the former to the latter (Fig
\ref{fig:fig5}; Pearl, 1988). However, the direction of the arc does not
necessarily imply causation, and the relationship between variables are
often described as probabilistic instead of casual (Scutari and Denis,
2014). The information within a node can be either continuous or
discrete, and a complete network can contain both continuous and
discrete nodes; however, discrete networks are the most commonly used
form of BN (Chen and Pollino, 2012). In discrete networks, conditional
probabilities for each child node are allocated for each combination of
the possible states in their parent nodes and can be used to assess the
strength of a dependency in the network.

\begin{figure}[H]
\caption{{\bf Example network.}
A simple discrete network contain nodes, possible states of the nodes and the arcs connecting nodes.}
\label{fig:fig5}
\end{figure}

In order to use discrete networks, continuous variables must first be
split into categorical levels. When there are a large number of
variables in the network, limiting the number of levels has the benefit
of producing a network that is more parsimonious in terms of parameters.
For example, a network with 10 variables each with two levels has 100
(10\^{}2) possible parameter combinations, however the same network with
three levels has 1000 (10\^{}3) possible parameter combinations, the
latter being significantly more computationally expensive. Using a
larger number of splits in the data also comes at a cost of reducing the
statistical power in detecting probabilistic associations, and reduces
the precision of parameter estimates for the probabilistic associations
that are detected because it reduces the sample-size-to-parameters ratio
(Scutari and Denis, 2014).

Learning the structure of the network is an important step in BN
modelling. The structure of a network can be constructed using expert
knowledge and/or data-driven algorithm techniques (e.g., search and
score, such as hill climbing and gradient descent algorithms; Scutari \&
Denis, 2014). The learned structure can then be used for inference by
querying the network\footnote{The term ``query'' in relation to Bayesian
  Networks stems from Pearl's expert systems theory (1988). A query can
  be submitted to an expert (in this case, the network is the expert) to
  get an opinion, the expert then updates the querier's beliefs
  accordingly. Widely used texts on Bayesian Network analysis (Koller
  and Friedman, 2009) have widely adopted the terminology in favour of
  that used in traditional statistics.} and obtaining the posterior
probabilities of a particular node for a given query. The posterior
distribution can be obtained by \(Pr(X|E,B) = Pr(X|E,G,\Theta)\), where
the learned network \(B\) with structure \(G\) and parameters
\(\Theta\), are investigated with new evidence \(E\) using the
information in \(B\) (Scutari and Denis, 2014). In the example network
presented in Fig \ref{fig:fig5}, new values assigned to each of the
parent nodes (e.g., both set to ``Low'') could be used to investigate
what effect the new information has on the state of the child node
(conditional probability of a particular state of the child node). In a
more complex network containing many nodes, the outcome of a particular
node can be assessed conditional on the states of any subset of nodes in
the network. BNs therefore provide a unique and versatile approach to
modelling a set of variables to uncover dependency structures within the
data.

BNs have recently been used in the sport psychology literature
(Olmedilla et al., 2018; Fuster-Parra et al., 2017) and offer several
benefits over traditional statistical analysis. For example, predictions
can be made about any variable in the network, rather than there being a
distinction between dependent and independent variables in the data,
such as in linear regression models that are often used within the sport
psychology literature (Olmedilla et al., 2018; Bittencourt et al.,
2016). Furthermore, the structure of a network can be obtained from both
empirical data \emph{and} prior knowledge about the area of study; the
latter being particularly useful when there are a large number of
variables in the network, or only a small number of observations are
available in the data (Xiao-xuan et al., 2008). In such instances, a
purely data driven approach to learning the network would be
time-consuming due to the large parameter space, and inefficient at
identifying an approximation of the true network structure. Prior
knowledge about dependencies between variables can therefore be included
in the network structure, while still allowing a data driven approach
for unknown dependencies, to improve the overall computation of the
network structure (Heckerman et al., 1995; Xu et al., 2015). The
following sections detail the steps taken in the current study to
firstly prepare the data for the network, and then obtain the structure
of the network that was used for inference.

\hypertarget{data-preparation}{%
\subsubsection*{Data Preparation}\label{data-preparation}}
\addcontentsline{toc}{subsubsection}{Data Preparation}

Of the 351 participants that were initially recruited for the study, 94
only completed the first time point, and therefore had to be removed
from the study as no injury information was obtained for these
participants following the first time point. To prepare the data for the
BN, missing values in the dataset were first imputed. Out of the 650
total measurements across all time points in the current study, there
were 31 (4.64\%) missing muscle stiffness measurements and 70 (10.48\%)
missing heart rate recordings. The missing data were due to technical
faults in the data collection equipment and were considered to be
missing completely at random. A missing rate of 15-20\% has been
reported to be common in psychological studies, and several techniques
are available to handle missing values (Enders, 2003; Lang et al.,
2013). In the current study, the \emph{caret} package (Kuhn et al.,
2008) was used to impute the missing values. A bagged tree model using
all of the non-missing data was first generated and then used to predict
each missing value in the dataset. The bagged tree method is a reliable
and accurate method for imputing missing values in data and is superior
to other commonly used methods such a median imputation (Kuhn et al.,
2008).

A median split technique was used to discretise the data used in the
network into ``Low'' and ``High'' levels. All variables apart from
negative and total life events were approximately normally distributed
(based on visual inspection, see supplementary table \ldots) and
required no further transformation prior to the median split. For the
LESCA questionnaire data, a cumulative total of the current, and
previous time points was calculated at each time point to account for
the potential continuing effect of the life events experienced by
athletes over time. Given the limited support for a relationship between
positive life events and injury (Williams and Andersen, 2007), only
negative and total life events were included in the network. Cumulative
negative, and cumulative total life event scores at each time point were
first log scaled so distributions were approximately normal, and then
binarised using the median at each time point (nlelg and tlelg
respectively). In addition to the log scaled cumulative values, an
untransformed NLE score from the first time point was included as an
additional variable based on previous literature that indicates this
variable should have a strong relationship with injury outcome (Ivarsson
et al., 2017).

\hypertarget{network-structure}{%
\subsubsection*{Network structure}\label{network-structure}}
\addcontentsline{toc}{subsubsection}{Network structure}

To obtain the network structure, several steps were taken to ensure that
both a theoretically realistic network, and a network that was an
appropriate fit to the collected data, was used for inference. Prior
knowledge about the network structure was included by providing a list
of arcs that are always \emph{restricted} from being in the network
(blacklist), and a list of arcs that are always \emph{included} in the
network (whitelist). Additionally, there are several scoring functions
such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet
equivalent uniform (BDeu) that can be used to compare network structures
with certain nodes and arcs included or excluded (Scutari and Denis,
2014). To account for the repeated measures design and to maximise the
use of the data, pairs of complete cases (e.g., participants who
completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network
(2TBN) structure (Murphy, 2002). In the 2TBN, variables measured T2
could depend on variables measured at T1 (e.g., T1 \(\rightarrow\) T2)
and variables measured at T3 could depend on variables measured at T2
(e.g., T2 \(\rightarrow\) T3). However, arcs were blacklisted between T2
\(\rightarrow\) T1 and T3 \(\rightarrow\) T2 to preserve the order in
which data was collected. Variables were separated into two groups;
``explanatory'', for variables that did not change during the study
(e.g., gender), or ``independent'', for variables that were measured at
each time point and could vary during the study. Independent variable
names were suffixed with \_1 for time point T, and \_2 for time point
T+1 (e.g., T1\_1 \(\rightarrow\) T2\_2 and T2\_1 \(\rightarrow\) T3\_2).
Formatting the data in this way meant participants who completed T1 and
T2, but did not complete T3, could still be included in the analysis.
Table \ref{tab:table2} provides an example of the formatted data.
Participants 1 and 3 have complete data, and therefore have two rows of
data each representing variables from T1 \(\rightarrow\) T2 and T2
\(\rightarrow\) T3, respectively. Participant 2 did not complete the
final data collection at T3 and therefore only has one row of data
representing the variables collected at T1 and T2. In addition to the
blacklisted arcs between T2 \(\rightarrow\) T1 and T3 \(\rightarrow\)
T2, the direction of arcs was restricted between independent variables
and explanatory variables (e.g., independent \(\rightarrow\)
explanatory); however, arcs were not restricted between explanatory
\(\rightarrow\) independent variables. Finally, arc direction was
restricted between specific nodes within the explanatory variables. Arcs
from clevel \(\rightarrow\) gender, nlebase \(\rightarrow\) gender and
nlebase \(\rightarrow\) ind\_team were included in the blacklist, as
arcs in these directions did not make logical sense. All subsequent
models used the same blacklist.

\begin{table}[H]

\caption{\label{tab:table2}Example of the data arrangement used for the network.}
\centering
\begin{tabular}[t]{l|l|l|l}
\hline
\textbf{Participant} & \textbf{X\_1} & \textbf{ } & \textbf{X\_2}\\
\hline
1 & T1 & -> & T2\\
\hline
1 & T2 & -> & T3\\
\hline
2 & T1 & -> & T2\\
\hline
3 & T1 & -> & T2\\
\hline
3 & T2 & -> & T3\\
\hline
\end{tabular}
\end{table}

\hypertarget{preliminary-network-structures}{%
\subsubsection*{Preliminary network
structures}\label{preliminary-network-structures}}
\addcontentsline{toc}{subsubsection}{Preliminary network structures}

Prior to the final network structure presented in the results section,
several structures with different combinations of variables were
explred. Networks were learned using a Tabu search algorithm (Russell
and Norvig, 2009) and BIC was used to compare different models. A higher
BIC value indicates the structure of a DAG is a better fit to the
observed data (Scutari and Denis, 2014). BIC values for each combination
of variables of interest are reported as the combination of variables
with the highest BIC value, followed by the relative scores of the other
variables in the model.

Initially, both negative life events and total life events were included
in the network structure, however, the network score was improved when
only nlelg or tlelg was included (highest BIC value = nleleg, BIC values
relative to nlelg; tlelg only = -85.13, tlelg and nlelg = -219.14).
Additionally, despite strong evidence in the literature that both
negative and total life event stress are related to injury occurrence
(Williams and Andersen, 2007), network structures learned using the Tabu
search algorithm failed to identify a relationship between NLE and
injury or TLE and injury in the data. Given that nlelg provided the
highest network score, and there is a stronger relationship between
negative life events and injury in the literature, an arc was
whitelisted between nlelg\_1 and injured\_1 and nlelg\_2 and injured\_2
in the final network structure. Total life event score was not included
in the final structure.

The subscales representing the BAS (RR, RI, GDP and I) showed limited
connection to other variables in the network. Therefore, several models
were run with each scale individually to find the scale that resulted in
the highest BIC value (values are shown relative to the highest value).
RI provided the highest BIC value, compared to RR (-13.13), GDP (-15.06)
and I (-15.2). Including all the variables (RR, RI, GDP and I) resulted
in a significantly lower score -894.37) indicating that including all
the variables was not beneficial to the model structure and did not
offset the cost of the additional parameters. Therefore, only RI was
included in the final structure.

Finally, both total score and asymmetry were included in the initial
network. However, visual inspection of the network revealed no arcs
between bal\_asym\_1 or bal\_asym\_2 and any other node in the network.
Therefore, balance asymmetry was removed from the final network
structure. To summarise, Table \ref{tab:table3} includes the variables
that were included in the final network structure.

\begin{table}[H]
\begin{adjustwidth}{-2.25in}{0in}
\caption{\label{tab:table3}Variables included in the final network structure.}
\centering
\begin{tabular}{l|l|l|l}
\hline
{\bf Variable} & {\bf Definition} & {\bf State} 1 & {\bf State 2}\\
\hline
clevel & Current competitive level & Club\_university\_county & National\_international\\
\hline
gender & Gender of the participant & Female & Male\\
\hline
hours & Number of hours spent training per week & 0-9 (Low) & >9-35 (High)\\
\hline
ind\_team & Participate in an individual or team based sport & Individual & Team\\
\hline
pi & Sustained previous injury & No Injury & Injury\\
\hline
nlebase & Untransformed NLE at TP 1 & 0-13 (Low) & >13-93 (High)\\
\hline
FFFS & Fight-Flight-Freeze System & 8-16 (Low) & >16-30 (High)\\
\hline
BIS & Behavioural Inhibition System & 17-38 (Low) & >38-68 (High)\\
\hline
RI & Reward Interest & 4-10 (Low) & >10-16 (High)\\
\hline
stiffness & Sum of all stiffness locations & 1543-2330 (Low) & >2330-4518 (High)\\
\hline
rmssd & Root mean squared difference of successive RR intervals & 2.03-4.02 (Low) & >4.02-5.94 (High)\\
\hline
balance & Total balance score & 5-15 (Low) & >15-46 (High)\\
\hline
nlelg\_1 & Log NLE at TP 1 & 0-2.64 (Low) & >2.64-4.54 (High)\\
\hline
nlelg\_2 & Log NLE at TP 2 & 0-3.04 (Low) & >3.04-5.19 (High)\\
\hline
nlelg\_3 & Log NLE at TP 3 & 0-3.18 (Low) & >3.18-4.79 (High)\\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}
